On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages\\ \large{Not For Distribution}

### 2020-07-30 paper/paper.tex
Better section references

\begin{itemize}
\item A representation of pattern matching in a Haskell EDSL (Section 3)
\item A representation of lambdas (Section 5) in the EDSL, taking advantage of
  the fact that both pattern matching and tail recursion are already captured in
  the EDSL.
\item A GHC Core plugin which transforms a subset of standard Haskell code
  into this EDSL for pattern matching, extended with tail recursion,
  lambdas and primitive operations (Section 8)
\item An implementation of a interpreter for the "standard" semantics of the EDSL (Section 6).
\item An outline of a C backend for this EDSL (Section 7).
\end{itemize}

Readers who are primarily interested only in the encoding of pattern matching itself can direct
most of their attention to Sections 2, 3 and 6. Most of the technical aspects of
the transformation performed by the Core plugin are described in Section 8.

@@

\section{Representing Algebraic Datatypes}

@@

There are three interconnected foundational parts: \ttt{E}, \ttt{ERep} and
\ttt{ERepTy}. \ttt{E} is the deep embedding of the EDSL (a GADT that encodes
expressions in the DSL language). \ttt{ERep} is a type class which represents
all Haskell types which can be represented in the DSL. \ttt{ERepTy} is a type
family associated to the \ttt{ERep} type class, which represents a "canonical form"
of the given type. This canonical form can be immediately constructed in the EDSL.
Canonical form types crucially include \ttt{Either} and \ttt{(,)}, which
allow all combinations of basic sum types and product types to be encoded, in the
manner described at the beginning of Section 2.

@@

\section{Representing pattern matches}

@@

\subsection{An aside on \ttt{ProdMatch} and \ttt{SumMatch} values}

@@

\section{Representing lambdas}
This representation of pattern matching depends on the ability to bring
function values into the expression language. This is accomplished
with the following constructors:

@@

\section{Recovering standard semantics for pattern matches}

The standard semantics for the EDSL is given by \ttt{abs}. Two helper
functions, \ttt{sumMatchAbs} and \ttt{prodMatchAbs}, are used to provide
the standard semantics for matches on sum types and matches on product
types, respectively. These helper functions are based on the mechanism
described in section 3.4.

@@

\section{C Backend}
\begin{todo}
  Should this section be expanded?
\end{todo}

@@

\section{Core Plugin}

@@

If an expression already has an EDSL type (a type of the form \ttt{E a} for some
\ttt{a}), then the marking procedure ignores it and does not wrap it with
\ttt{externalize} (see $M$ in Section 8.2).

@@

Elliott developed GHC plugins~\cite{github:lambda-ccc}\cite{github:reification-rules}
for compiling Haskell to hardware~\cite{github:Elliott:Talk:2015},
using worker-wrapper style transformations~\cite{Gill:09:WW}
equivalent to the \verb|abs| and
\verb|rep| transformations described in Section \ref{sec:Overview}. These plugins were later generalized to enable additional interpretations~\cite{Elliott:2017}.

>>>>

\begin{itemize}
  \item A representation of pattern matching in a Haskell EDSL (Section~\ref{sec:PatRep})
  \item A representation of lambdas (Section~\ref{sec:LamRep}) in the EDSL, taking advantage of
  the fact that both pattern matching and tail recursion are already captured in
  the EDSL.
\item A GHC Core plugin which transforms a subset of standard Haskell code
  into this EDSL for pattern matching, extended with tail recursion,
  lambdas and primitive operations (Section~\ref{sec:CorePlugin})
\item An implementation of a interpreter for the "standard" semantics of the EDSL (Section~\ref{sec:StdSemantics}).
\item An outline of a C backend for this EDSL (Section~\ref{sec:CBackend}).
\end{itemize}

Readers who are primarily interested only in the encoding of pattern matching itself can direct
most of their attention to Sections~\ref{sec:ADTRep},~\ref{sec:PatRep} and~\ref{sec:StdSemantics}. Most of the technical aspects of
the transformation performed by the Core plugin are described in Section~\ref{sec:CorePlugin}.

@@

\section{Representing Algebraic Datatypes}
\label{sec:ADTRep}

@@

There are three interconnected foundational parts: \ttt{E}, \ttt{ERep} and
\ttt{ERepTy}. \ttt{E} is the deep embedding of the EDSL (a GADT that encodes
expressions in the DSL language). \ttt{ERep} is a type class which represents
all Haskell types which can be represented in the DSL. \ttt{ERepTy} is a type
family associated to the \ttt{ERep} type class, which represents a "canonical form"
of the given type. This canonical form can be immediately constructed in the EDSL.
Canonical form types crucially include \ttt{Either} and \ttt{(,)}, which
allow all combinations of basic sum types and product types to be encoded, in the
manner described at the beginning of Section~\ref{sec:ADTRep}.

@@

\section{Representing pattern matches}
\label{sec:PatRep}

@@

\subsection{An aside on \ttt{ProdMatch} and \ttt{SumMatch} values}
\label{sec:MatchAside}

@@

\section{Representing lambdas}
\label{sec:LamRep}

This representation of pattern matching depends on the ability to bring
function values into the expression language. This is accomplished
with the following constructors:

@@

\section{Recovering standard semantics for pattern matches}
\label{sec:StdSemantics}

The standard semantics for the EDSL is given by \ttt{abs}. Two helper
functions, \ttt{sumMatchAbs} and \ttt{prodMatchAbs}, are used to provide
the standard semantics for matches on sum types and matches on product
types, respectively. These helper functions are based on the mechanism
described in Section~\ref{sec:MatchAside}.

@@

\section{C Backend}
\label{sec:CBackend}

\begin{todo}
  Should this section be expanded?
\end{todo}

@@

\section{Core Plugin}
\label{sec:CorePlugin}

@@

If an expression already has an EDSL type (a type of the form \ttt{E a} for some
\ttt{a}), then the marking procedure ignores it and does not wrap it with
\ttt{externalize} (see $M$ in Figure~\ref{fig:Rewrites}).

@@

Elliott developed GHC plugins~\cite{github:lambda-ccc}\cite{github:reification-rules}
for compiling Haskell to hardware~\cite{github:Elliott:Talk:2015},
using worker-wrapper style transformations~\cite{Gill:09:WW}
equivalent to the \verb|abs| and
\verb|rep| transformations described in Section~\ref{sec:Overview}. These plugins were later generalized to enable additional interpretations~\cite{Elliott:2017}.
